---
title: "Stat 696 Report Template"
author: "Your Name"
date: "The date"
# header-includes:
#   - \usepackage[title]{appendix}
output: 
  bookdown::pdf_document2:
    toc: false
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Executive Summary {-}

this is my summary


# Introduction


Note: you need to install the R package `bookdown` for this to work.

If you want the Appendices to be labeled "Appendix A", then you can remove the comments in the header (just the "#", not any spaces) and the end of the document.

  




# Analysis

blah blah blah blah blah blah blah blah blah

## Exploratory Analysis

The data is in Table \@ref(tab:cars-data).

```{r cars-data}
knitr::kable(head(mtcars),
             caption = "The data.")
```


blah blah blah blah blah blah blah blah blah

Figure \@ref(fig:cars-plot) is a box plot

```{r cars-plot, fig.cap = "A box plot."}
boxplot(mpg ~ cyl, data = mtcars)
```



## Modeling and Diagnostics

blah blah blah blah blah blah blah blah blah


# Conclusion



\newpage

# Appendix {-}

# (APPENDIX) Appendix A {-}


# Supplementary Figures and Model Stuff

## Figures

## Model Stuff


# R Code


<!--
\begin{appendices}

\section{Supplementary Figures}

\section{Model Stuff}

\end{appendices}
-->


