---
title: "Something About Prostate Cancer"
author: "Andrew Bates"
date: "November 15, 2018"
header-includes:
  - \renewcommand{\abstractname}{Executive Summary}
output: 
  bookdown::pdf_document2:
    toc: false
documentclass: article
geometry: margin=1in
fontsize: 11pt
abstract: "This is the executive summary. Unfortunately, it has to be typed in the .yaml header. The only other thing i can think of is to have a separate file for the abstract. I'm not sure I want to do this. But actually, this may not be so bad. Each section could have a different file. this might make things a bit easier to edit. because then i would only have to look at say, the conclusion file instead of having to scroll all the way down, passing it, srolling up, passing it, etc."
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(here)
library(readr)
library(dplyr)

prostate <- read_tsv(here("data", "prostate.txt"))
```





# Introduction

In this paper, we investigate the relationship between the results of various prostae related exams and whether tumor penetration of the prostatic capsule has occured. We examine a subset of data collected by the Ohio State University Comprehensive Cancer Center as part of a study to determine the potential of standard exam results to predict whether a tumor will penetrate the prostatic capsule. Out of $380$ patients $153$ have experienced capsule penetration and $227$ have not. The data set contains six explanatory variables: the patients age and race (Black or White), results of a digital rectal exam, whether capsular involvement was detected, Prostatic Specif Antigen value, and total Gleason score.

<!-- Table: (\#tab:var-description) Description of variables in the data set. -->

<!-- | Variable Name | Description | Details | -->
<!-- |:------|:-----|:-----| -->
<!-- | capsule | Did the tumor penetrate the prostatic capsule | Yes, no | -->
<!-- | age | Patient age| In years | -->
<!-- | race | Patient race | Black or White | -->
<!-- | dre | Results of digital rectal exam | No nodule, unilobar nodule left, unilobar nodule left right, bilobar nodule | -->
<!-- | caps | Detection of capsular involvement | Yes, no | -->
<!-- | psa | Prostatic Specific Antigen value | mg/ml | -->
<!-- | gleason | Total Gleason score | 0-10 | -->


Table \@ref(tab:var-desc) contains a detailed description of the variables in this data set.

```{r var-desc, results = "asis"}
description <- data.frame(
  Name = c("capsule", "age", "race", "dre", "caps", "psa", "gleason"),
  Description = c("Tumor penetration of prostatic capsule?", "Patient age", "Patient race", "Results of digital rectal exam", "Detection of capsular involvement?", "Prostatic Specific Antigen value", "Total Gleason score"),
  Details = c("Yes, no", "Years", "Black, White", "No nodule, unilobar left, unilobar right, bilobar", "Yes, no", "mg/ml", "0-10")
)

knitr::kable(description,
             caption = "Description of variables in the data set.")

```



# Analysis

blah blah blah blah blah blah blah blah blah

## Exploratory Analysis



## Modeling and Diagnostics

blah blah blah blah blah blah blah blah blah


# Conclusion



\newpage

# Appendix {-}

# (APPENDIX) Appendix A {-}


# Supplementary Figures and Model Stuff

## Figures

## Model Stuff


# R Code




