---
title: "Something About Prostate Cancer"
author: "Andrew Bates"
date: "November 15, 2018"
output: 
  bookdown::pdf_document2:
    toc: false
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Executive Summary {-}

this is my summary


# Introduction

In this paper, we investigate the relationship between the results of various prostae related exams along with individual factors and whether tumor penetration of the prostatic capsule has occured.

The data at hand is a subset of data collected by the Ohio State University Comprehensive Cancer Center as part of a study to determine the potential of standard exam results in predicting whether a tumor has penetrated the prostatic capsule.
  




# Analysis

blah blah blah blah blah blah blah blah blah

## Exploratory Analysis

The data is in Table \@ref(tab:cars-data).

```{r cars-data}
knitr::kable(head(mtcars),
             caption = "The data.")
```


blah blah blah blah blah blah blah blah blah

Figure \@ref(fig:cars-plot) is a box plot

```{r cars-plot, fig.cap = "A box plot."}
boxplot(mpg ~ cyl, data = mtcars)
```



## Modeling and Diagnostics

blah blah blah blah blah blah blah blah blah


# Conclusion



\newpage

# Appendix {-}

# (APPENDIX) Appendix A {-}


# Supplementary Figures and Model Stuff

## Figures

## Model Stuff


# R Code




